Molecular dynamics simulation of synchronization in driven particles

### 2021-01-04 manuscript/manuscript_ajp_2020.tex
finished simple Reynolds number exercise

%because 
We model the 
particle dynamics 
with an overdamped
equation of motion.
%integrated with the Verlet method.
%The suspending fluid is highly viscous
%and exerts a damping force on the particles equivalent
%to a linear drag force.
%Overdamped dynamics is a common assumption in models of colloidal particles.
%This is an assumption that
%The term overdamped or critically damped indicates
%that an oscillator cannot undergo periodic
%undulations due to some external force. %such as
%suspension in a viscous fluid \cite{}.
In our colloidal model we assume
the damping comes from the viscosity of the fluid,
where a sphere of radius $0.5 \mu m$ has a Reynolds number
of $R \approx 10^{-6}$, indicating that linear drag dominates
the particle motion.
See Section~\ref{sec:problems} Exercise~\ref{ex:reynolds} for more information
about the Reynold's number.
In the overdamped model the
linear friction produced by this 
nonconservative force,
$\vec{F}_{drag} = -\eta \vec{v}_i$ is 
sufficient so that
the acceleration of the particle is zero.
Such a model is appropriate when the particles
are small and the viscosity of the suspending medium is high.

Newton's second law for an individual particle
$m \vec{a}_i = \sum \vec{F}$ is simplified
by the assumption $\vec{a}_i$ is zero in overdamped model.
%Newton's second law can be rearranged %for the velocity
%to
The equation of motion for an isolated particle is
%solved for velocity 
\begin{equation}
  \eta \vec{v}_i = \vec{F}^l_{i} + \vec{F}^{D}(t).
    \label{eq:motion}
\end{equation}
with the viscosity coefficient $\eta = 1$,
which we describe in more detail in Exercise~\ref{ex:reynolds}.
In Section~\ref{sec:problems} Exercise~\ref{ex:n2l}
we invite the reader to confirm this result.

@@

\begin{table}[h!]
\centering
\caption{Simulation units.}
\begin{ruledtabular}
\begin{tabular}{c c p{5cm}}
Quantity & Unit \\
\hline
length &  $a_0 = 1$ \\
energy & $E_0 = 1$ \\
%electric potential & $V(r_{ij}) = E_0/r_{ij}$ \\
%energy & $E_0 = q^2{Z^*}^2/4\pi \epsilon \epsilon_0 a_0$ \\
%dimensionless interaction strength & $q$ \\
%effective colloidal charge & $Z^*$ \\
%solvent dielectric constant & $\epsilon \epsilon_0$\\
force & $E_0 / a_0$\\
viscosity coefficient & $\eta = 1$ \\
velocity &  $ E_0 / (\eta a_0) $ \\
time &  $(\eta / E_0) {a_0}^2$ \\
\end{tabular}
\end{ruledtabular}
\label{tab:1}
\end{table}

@@

%single DC case with figure
%dc_current 0.01
%ac_current 0.2
%ac_frequency 0.0001
%fp_y 0.1
\begin{center}
\begin{figure}[h!]
\centering
\includegraphics[width=\columnwidth]{single_particle.pdf}
\caption{(a) The applied driving force $F^D(t)$ (orange) where the parameter $F^{DC}$ (blue) is ramped to a constant value.  (b) 
  The position as a function of time of a single driven particle
  normalized by the period of the substrate $\lambda$.
  }
\label{fig:0}
\end{figure}
\end{center}

@@

\begin{enumerate}
\item Stokes' law the viscous drag on a sphere is
$\vec{F}_{lin} = 3 \pi \eta D \vec{v}$ \cite{}
where the fluid viscosity is $\eta$,
$D$ is the particle diameter.
Assuming a colloid with diameter 
$D = 1 \mu m$ and
the viscosity of water at room temperature is
$\eta = 0.01 cm^2/sec$ \cite{} %CONFIRM!
%density $\rho = 0.9982 g/cm^3$
%https://ascelibrary.org/doi/pdf/10.1061/9780784408230.ap02
Reynold's number $R = D v \rho / \eta$
where $\rho$ is the fluid density and $v$ the particle's speed.
When $R$ is small, the quadratic and higher order drag terms
may be ignored in favor of the linear drag term.
A velocity $v$ in experiment is on the order of $v \sim 1 \mu /s$,
thus $R \approx 10^-6$.

This model should be familiar to readers
of a standard classical mechanics text, such as Ref.~\cite{Taylor2005}
when studying 
models Millikan's oil drops.
\label{ex:reynolds}

@@

\item Brownian motion without fluid dynamics using  Ermak's algorithm

@@

\subsection{Grid Contour Plot of landscape}
%code is located in
%~/pymodules/animation_code/channel_colloid_movie_maker.py
Testing section - how will putting the code directly in the appendix appear?

@@

    \bibitem{Frenkel2001} D. Frenkel and B. Smit, Understanding Molecular Simulation: From Algorithms to Applications (Academic Press, London, 2001).
    \bibitem{Allen2017} M. P. Allen and D. J. Tildesley, Computer Simulation of Liquids.  Second Edition. Oxford University Press (2017).

@@

>>>>

%because 
We model the 
particle dynamics 
with an overdamped
equation of motion.
%integrated with the Verlet method.
%The suspending fluid is highly viscous
%and exerts a damping force on the particles equivalent
%to a linear drag force.
%Overdamped dynamics is a common assumption in models of colloidal particles.
%This is an assumption that
%The term overdamped or critically damped indicates
%that an oscillator cannot undergo periodic
%undulations due to some external force. %such as
%suspension in a viscous fluid \cite{}.
In our colloidal model we assume
the particle is suspending in a viscous fluid
that provides friction, or damping.
In the overdamped model the
linear friction produced by this 
nonconservative force,
$\vec{F}_{drag} = -\eta \vec{v}_i$ is
sufficient so that
the acceleration of the particle is zero.
The friction coefficient $\eta$
is proportional to the fluid viscosity.
Such a model in which inertia is set to zero
is appropriate for small particles
moving through a high viscosity fluid
such as a colloid moving in water.
We discuss this further in 
Section~\ref{sec:problems} Exercise~\ref{ex:reynolds}. %for more information

Newton's second law for an individual particle
$m \vec{a}_i = \sum \vec{F}$ is simplified
by the assumption $\vec{a}_i$ is zero in overdamped model.
%Newton's second law can be rearranged %for the velocity
%to
The equation of motion for an isolated particle is
%solved for velocity 
\begin{equation}
  \eta \vec{v}_i = \vec{F}^l_{i} + \vec{F}^{D}(t).
    \label{eq:motion}
\end{equation}
with the viscosity coefficient $\eta = 1$ in units,
which we describe in more detail in Exercise~\ref{ex:reynolds}.
In Section~\ref{sec:problems} Exercise~\ref{ex:n2l}
we invite the reader to confirm this result.

@@

\begin{table}[h!]
\centering
\caption{Simulation units.}
\begin{ruledtabular}
\begin{tabular}{c c p{5cm}}
Quantity & Unit \\
\hline
length &  $a_0 = 1$ \\
energy & $E_0 = 1$ \\
%electric potential & $V(r_{ij}) = E_0/r_{ij}$ \\
%energy & $E_0 = q^2{Z^*}^2/4\pi \epsilon \epsilon_0 a_0$ \\
%dimensionless interaction strength & $q$ \\
%effective colloidal charge & $Z^*$ \\
%solvent dielectric constant & $\epsilon \epsilon_0$\\
force & $F_0 = E_0 / a_0$\\
velocity &  $ a_0 / \tau $ \\
viscosity coefficient & $\eta = F_0 \tau / a_0 = 1 $\\
time &  $ \tau = \eta a_0 / F_0 $ \\

\end{tabular}
\end{ruledtabular}
\label{tab:1}
\end{table}

@@

%single DC case with figure
%dc_current 0.01
%ac_current 0.2
%ac_frequency 0.0001
%fp_y 0.1
\begin{center}
\begin{figure}[h!]
\centering
\includegraphics[width=\columnwidth]{single_particle.pdf}
\caption{(a) The applied driving force $F^D(t)$ (orange) in Eq.~\ref{eq:drive}
  where the parameter $F^{DC}$ (blue) is slowly increased from zero to a constant value
  as described in the text.  (b) 
  The $y-$position of the driven particle as a function of time %of a single driven particle
  normalized by the period of the substrate $\lambda$.
  }
\label{fig:0}
\end{figure}
\end{center}

@@

\begin{enumerate}
\item Stokes' law describes the viscous drag on a sphere
  moving at velocity $\vec{v}$ as 
  $\vec{F}^{lin} = 3 \pi \eta D \vec{v}$ \cite{}
  where $\eta$ is the dynamic fluid viscosity and 
  $D$ is the particle diameter.
  (In simulation we
  subsume the constants $3 \pi D$
  such that $\eta \rightarrow 3 \pi D \eta$.)
  Often drag forces are
  modeled as a polynomial series
  $\vec{F}^{drag} = \vec{F}^{lin} + \vec{F}^{quad} + ... $.
  Truncating the series to the first term
  is justified by calculating the Reynolds number  
  $R = D v \rho / \eta$
  where $\rho$ is the fluid density and $v$ the particle's speed.
  When $R$ is small, the quadratic and higher order drag terms
  may be ignored in favor of the linear drag term.

  Demonstrate the Reynolds number is low
  for this system.

  Reasonable experimental values for
  a colloid are $D = 1 \mu m$ and $v \sim 1 \mu m /s$.
  Viscosity of water decreases with temperature
  from $1.3 \times 10^{-3} Pa s$ at $10^{\circ}C$ to 
  from $0.31 \times 10^{-3} Pa s$ at $90^{\circ}C$,
  making $\eta \sim 10^{-3} Pa s $.
  Likewise the density $\rho$ varies with temperature
  and $\rho \sim 10^3 kg/cm^3$ is a reasonable approximation.

  [citation for these values better than wikipedia] \cite{} %CONFIRM!
  %density $\rho = 0.9982 g/cm^3$
  %https://ascelibrary.org/doi/pdf/10.1061/9780784408230.ap02

This model should be familiar to readers
of a standard classical mechanics text, such as Ref.~\cite{Taylor2005}
when studying 
models Millikan's oil drops.

[could certainly do more with this question to make it more computational.  For instance, plot $R$ as a function of any of the parameters $v$, $D$ and $\rho$, include a temperature dependent $\eta(T)$ in the Brownian system.  Suggest students model something big and include inertia and quadratic drag, but then we're fairly off topic...]

\label{ex:reynolds}

@@

\item Brownian motion.  Consider the consequences of a non-zero temperature on the simulation, in which particles will collide with invisible particles making up the suspending fluid.  These collisions are more likely at higher temperature due to the increased kinetic energy of the fluid particles, as described by the Maxwell-Boltzmann distribution \cite{}.  In molecular dynamics simulations temperature effects can be modeled by including randomized forces $f^T$ to emulate the motion of particles undergoing Brownian motion.  A distribution of randomizes forces contributes equally in all directions such that the force averaged over a finite time interval is zero $\langle f^T(t) \rangle = 0$.  Random distributions of contain no correlations over the independent variable (time in this case) $\langle f^T(t) f^T(\tau)\rangle = 2 \eta k_B T \delta(t-\tau)$ where the energy $k_B T$ derived from the Boltzmann constant $k_B$ and temperature $T$ has energy units $E_0$.  \cite{Allen2017}

  With no applied driving force, measure the minimum temperature required for a single particle to hop between minima in the potential landscape.

  Measure the rate of diffusion.

@@

@@

    \bibitem{Frenkel2001} D. Frenkel and B. Smit, Understanding Molecular Simulation: From Algorithms to Applications (Academic Press, London, 2001).

    \bibitem{Allen2017} M. P. Allen and D. J. Tildesley, Computer Simulation of Liquids.  Second Edition. Oxford University Press (2017).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Grid Contour Plot of landscape}
%code is located in
%~/pymodules/animation_code/channel_colloid_movie_maker.py
Testing section - how will putting the code directly in the appendix appear?
